---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

<style>
.caption {
    color: #3399ff;		
	font-weight: bold;
		}
</style>

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  res = 140, 
  fig.path = "man/figures/README-",
  out.width = "50%"
)
```

```{r setup, message = FALSE, include = FALSE}
library(ratioScales)
library(tidyverse)
library(captioner)
# tab_cap <- captioner(prefix = "Table")
# fig_cap <- captioner(prefix = "Figure")
theme_set(theme_classic() +
          theme(axis.text.x = element_text(angle = 45,
                                           vjust = 1, hjust = 1)
                , text = element_text(size = 14)))

```

# ratioScales

<!-- badges: start -->

[![R-CMD-check](https://github.com/mikeroswell/ratioScales/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/mikeroswell/ratioScales/actions/workflows/R-CMD-check.yaml)
[![CRAN status](https://www.r-pkg.org/badges/version/ratioScales)](https://CRAN.R-project.org/package=ratioScales)
<!-- badges: end -->

Logarithmic axis scales can clearly communicate multiplicative changes; they can 
also confuse. **ratioScales** annotates logarithmic axis scales with tickmarks 
that denote proportional and multiplicative change simply and explicitly.  

The main function in this package, `scale_*_ratio()`, is ggplot-friendly and 
works similarly to existing`scale_*_*` functions from **ggplot2** and 
**scales**. 

## Installation

You can install the development version of ratioScales from 
[GitHub](https://github.com/) with:

``` r
# install.packages("remotes")
remotes::install_github("mikeroswell/ratioScales")
```

## Example

Consider exchange rates between US and Canadian dollars:

```{r raw_exchange, warning = FALSE, fig.cap = "Exchange rates between US and Canada during the COVID-19 pandemic", fig.alt = "Exchange rates between US and Canada during the COVID-19 pandemic. x-axis: dates from April 2020 to April 2022. y-axis: exchange rate (Canadian to US in Blue, US to Canadian in Magenta). Two curves with imperfect mirroring on arithmetic scale."}

exch %>% 
  ggplot(aes(date, exRate, color = direction)) + 
  geom_point() +
  scale_color_manual(values = hcl.colors(4, "Plasma")[c(1,2)]) +
  labs(y = "exchange rate") 
```

```{r first fig cap, echo = FALSE}
# fig_cap("fig1", "Exchange rates between US and Canada during the COVID-19 Pandemic")
```

Let's see, relative to some baseline (1 April 2020), is the Canadian dollar
gaining or losing ground against the US dollar, and by how much?

```{r scaled_exchange, warning = FALSE, fig.cap = "Proportional change in exchange rate from 1 April 2020 to 1 April 2022", fig.alt = "add alt text here"}

 exch %>%  ggplot(aes(date, exRate_scale, color = direction)) + 
   geom_hline(yintercept = 1, color = "black")+
   geom_point() +
   scale_color_manual(values = hcl.colors(4, "Plasma")[c(1,2)]) +
   geom_hline(yintercept = 0.85
              , color = hcl.colors(4, "Plasma")[1], linetype = 3) +
   geom_hline(yintercept = 1.15
              , color = hcl.colors(4, "Plasma")[2], linetype = 5) +
  # FUNCTION FROM ggplot2
   scale_y_continuous(breaks = seq(80, 130, 5)/100) +
   labs(y = "proportional change in exchange rate") 
```

But this is strange! Somehow the Canadian dollar *weakened* by 
**a maximum of 15%** before rebounding, but the US dollar *strengthened* by 
**much more than 15%**. Maybe not the best way to think about this? 

**ratioScales** provides "rational" alternatives.

### divMult scale

The "divMult" scale in **ratioScales** shows absolute ratios, prefaced by an
operator sign (e.g., ร or รท), allowing easy and accurate comparison of
multiplicative changes.

```{r divMult_example, warning = FALSE, fig.cap = "exchange rate changes on the divMult scale", fig.alt = "Add alt text"}
exch %>%  ggplot(aes(date, exRate_scale, color = direction)) + 
   geom_hline(yintercept = 1, color = "black")+
  # times and divided by 1.15; longdash
   geom_hline(yintercept = 1/1.15
              , color = hcl.colors(4, "Plasma")[1]
              , linetype = 5) +
   geom_hline(yintercept = 1.15
              , color = hcl.colors(4, "Plasma")[2]
              , linetype = 5) +
  # times and divided by 0.85; dotted
   geom_hline(yintercept = 1/0.85
              , color = hcl.colors(4, "Plasma")[2]
              , linetype = 3) +
   geom_hline(yintercept = 0.85
              , color = hcl.colors(4, "Plasma")[1]
              , linetype = 3) +
   geom_point() +
   scale_color_manual(values = hcl.colors(4, "Plasma")[c(1,2)]) +
  # FUNCTION FROM ratioScales
   scale_y_ratio(tickVal = "divMult", n = 12, nmin = 12, slashStar = FALSE) +
   labs(y = "multiplicative change in exchange rate") 
```

### percDiff

Prefer percentage differences? You can do this in a principled fashion using
`scale_y_ratio(tickVal = "percDiff")`.

```{r percDiff_example, warning = FALSE, fig.cap = "caption explaining the reference lines"}
exch %>%  ggplot(aes(date, exRate_scale, color = direction)) + 
   geom_hline(yintercept = 1, color = "black")+
  # times and divided by 1.15; longdash
   geom_hline(yintercept = 1/1.15
              , color = hcl.colors(4, "Plasma")[1]
              , linetype = 5) +
   geom_hline(yintercept = 1.15
              , color = hcl.colors(4, "Plasma")[2]
              , linetype = 5) +
  # times and divided by 0.85; dotted
   geom_hline(yintercept = 1/0.85
              , color = hcl.colors(4, "Plasma")[2]
              , linetype = 3) +
   geom_hline(yintercept = 0.85
              , color = hcl.colors(4, "Plasma")[1]
              , linetype = 3) +
   geom_point() +
   scale_color_manual(values = hcl.colors(4, "Plasma")[c(1,2)]) +
    # FUNCTION FROM ratioScales
   scale_y_ratio(tickVal = "percDiff") +
   labs(y = "percentage difference in exchange rate") 
```

This preserves the true ratio-based differences on the visual plot, but the values on the guide do not correspond simply to ratio differences (and are not symmetric, see plot).

### centiNels

<metatext>
We can also make a plot that will often be very similar to the percDiff plot, but which uses numbers that are better for quantitative analysis.
</metatext>

<!-- some COMMENTS here to keep track of 
We think we are currently using GitHub Actions to re-render `README.Rmd` every time you push. An example workflow can be found here: <https://github.com/r-lib/actions/tree/v1/examples>.

When you add a new plot, don't forget to commit and push the resulting figure files, so they display on GitHub and CRAN.

Add alt-text (and consider adding short captions as well) to all figures
-->
