---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

<style>


```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  res = 140, 
  fig.path = "man/figures/README-",
  out.width = "50%"
)
```

```{r setup, message = FALSE, include = FALSE}
library(ratioScales)
library(tidyverse)
```

# ratioScales

<!-- badges: start -->

[![R-CMD-check](https://github.com/mikeroswell/ratioScales/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/mikeroswell/ratioScales/actions/workflows/R-CMD-check.yaml)
[![CRAN status](https://www.r-pkg.org/badges/version/ratioScales)](https://CRAN.R-project.org/package=ratioScales)
<!-- badges: end -->

Logarithmic axis scales can clearly communicate multiplicative changes; they can 
also confuse. **ratioScales** annotates logarithmic axis scales with tickmarks 
that denote proportional and multiplicative change simply and explicitly.  

The main function in this package, `scale_*_ratio()`, is ggplot-friendly and 
works similarly to existing`scale_*_*` functions from **ggplot2** and 
**scales**. To learn more, please visit [https://mikeroswell.github.io/ratioScales/](https://mikeroswell.github.io/ratioScales/)

## Installation

You can install the development version of ratioScales from 
[GitHub](https://github.com/) with:

``` r
# install.packages("remotes")
remotes::install_github("mikeroswell/ratioScales")
```

