---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  res = 140, 
  fig.path = "man/figures/README-",
  out.width = "50%"
  
)
```
```{r setup, message = FALSE, include = FALSE}
library(ratioScales)
library(tidyverse)
theme_set(theme_classic() +
          theme(axis.text.x = element_text(angle = 45,
                                           vjust = 1, hjust = 1)
                , text = element_text(size = 14)))

```

# ratioScales

<!-- badges: start -->

[![R-CMD-check](https://github.com/mikeroswell/ratioScales/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/mikeroswell/ratioScales/actions/workflows/R-CMD-check.yaml)
[![CRAN status](https://www.r-pkg.org/badges/version/ratioScales)](https://CRAN.R-project.org/package=ratioScales)
<!-- badges: end -->

Logarithmic axis scales can clearly communicate multiplicative changes; they can 
also confuse. **ratioScales** annotates logarithmic axis scales with tickmarks 
that denote proportional and multiplicative change simply and explicitly.  

## Installation

You can install the development version of ratioScales from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("mikeroswell/ratioScales")
```

## Example

Consider exchange rates between US and Canadian dollars:

```{r raw_exchange, warning = FALSE}

exch %>% 
  ggplot(aes(date, exRate, color = direction)) + 
  geom_point() +
  scale_color_manual(values = hcl.colors(4, "Plasma")[c(1,2)]) +
  labs(y = "exchange rate") 
```

It might make more sense to ask relative to some baseline (1 April 2020), is the 
Canadian dollar gaining or losing ground against the US dollar, and by how much?

```{r scaled_exchange, warning = FALSE}

 exch %>%  ggplot(aes(date, exRate_scale, color = direction)) + 
   geom_hline(yintercept = 1, color = "black")+
   geom_point() +
   scale_color_manual(values = hcl.colors(4, "Plasma")[c(1,2)]) +
   geom_hline(yintercept = 0.85, color = hcl.colors(4, "Plasma")[1], linetype = 3) +
   geom_hline(yintercept = 1.15, color = hcl.colors(4, "Plasma")[2], linetype = 5) +
   scale_y_continuous(breaks = seq(80, 130, 5)/100) +
   labs(y = "proportional change in exchange rate") 
```

But this is strange! Somehow the Canadian dollar *weakend* by _a maximum of 15%_ 
before rebounding, but the US dollar *strengthened* by _much more than 15%_. 
Maybe not the best way to think about this? **ratioScales** provides "rational" 
alternatives. For example, using a "divMult" scale:

```{r divMult_example, warning = FALSE}
exch %>%  ggplot(aes(date, exRate_scale, color = direction)) + 
   geom_hline(yintercept = 1, color = "black")+
  # times and divided by 1.15; longdash
   geom_hline(yintercept = 1/1.15, color = hcl.colors(4, "Plasma")[1], linetype = 5) +
   geom_hline(yintercept = 1.15, color = hcl.colors(4, "Plasma")[2], linetype = 5) +
  # times and divided by 0.85; dotted
   geom_hline(yintercept = 1/0.85, color = hcl.colors(4, "Plasma")[2], linetype = 3) +
   geom_hline(yintercept = 0.85, color = hcl.colors(4, "Plasma")[1], linetype = 3) +
   geom_point() +
   scale_color_manual(values = hcl.colors(4, "Plasma")[c(1,2)]) +
   scale_y_ratio(tickVal = "divMult", n = 12, nmin = 12, slashStar = FALSE) +
   labs(y = "multiplicative change in exchange rate") 
```

Stuck on percent differences? That's ok too, if you use an appropriate scale:

```{r percDiff_example, warning = FALSE}
exch %>%  ggplot(aes(date, exRate_scale, color = direction)) + 
   geom_hline(yintercept = 1, color = "black")+
  # times and divided by 1.15; longdash
   geom_hline(yintercept = 1/1.15, color = hcl.colors(4, "Plasma")[1], linetype = 5) +
   geom_hline(yintercept = 1.15, color = hcl.colors(4, "Plasma")[2], linetype = 5) +
  # times and divided by 0.85; dotted
   geom_hline(yintercept = 1/0.85, color = hcl.colors(4, "Plasma")[2], linetype = 3) +
   geom_hline(yintercept = 0.85, color = hcl.colors(4, "Plasma")[1], linetype = 3) +
   geom_point() +
   scale_color_manual(values = hcl.colors(4, "Plasma")[c(1,2)]) +
   scale_y_ratio(tickVal = "percDiff") +
   labs(y = "percentage difference in exchange rate") 
```


<!-- some comments here to keep track of 
You'll still need to render `README.Rmd` regularly, to keep `README.md` up-to-date. `devtools::build_readme()` is handy for this. You could also use GitHub Actions to re-render `README.Rmd` every time you push. An example workflow can be found here: <https://github.com/r-lib/actions/tree/v1/examples>.

You can also embed plots, for example:

```{r pressure, echo = FALSE}
plot(pressure)
```

In that case, don't forget to commit and push the resulting figure files, so they display on GitHub and CRAN.
-->
