---
title: "centiNels"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{centiNels}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
  , fig.width = 7
  , fig.height = 7
)
```

```{r setup, message=FALSE}
library(ratioScales)
library(tidyverse)
theme_set(theme_classic() +
          theme(axis.text.x = element_text(angle = 45,
                                           vjust = 1, hjust = 1)))
```

On 1 April, 2020, a Canadian dollar (CAD) cost 0.703 US dollar (USD). The
Canadian dollar gained strength pretty steadily for the next year, and by 1 July
it would cost 0.736 USD, by  1 October, 0.753 USD, by 4 January 2021, 0.782 USD,
and by April fools day 2021, 0.796 USD. How much is the exchange rate changing?

Often, such changes are described in terms of percent difference, even though
this is sort of insane. Below is a brief peak at the insanity of percentage
differences: any change in the exchange rate will be reflected as different
percentage changes, depending on which direction you're hoping to exchange
currency:

```{r percents are insane}
exch %>% 
  filter(date %in% as.Date(c("2020-04-01"
                     , "2020-07-01"
                     , "2020-10-01"
                     , "2021-01-04"
                     , "2021-04-01"))) %>% 
  mutate(percent_difference = round(100*(exRate_scale-1), 2)) %>% 
  select(date, direction, exRate, percent_difference) %>% 
  pivot_wider(id_cols = c("date")
              , names_from = "direction"
              , values_from = c("exRate", "percent_difference")
              , names_vary = "slowest"
              )
```

On 1 April 2021, US dollars had gotten much cheaper for Canadians than they had
been a year previously... one US dollar cost less than 90% as much as a it had a
year earlier (or a US dollar had lost over 10% of its value in CAD). Amazingly,
over the same period, Canadian dollars got over 11% more expensive for
Americans. 

To put this more concretely, if Jonathan traded 100 CAD with Michael for 70.31
USD on 1 April 2020, doubtless he'd feel the fool come 1 April, 2021, when
Michael would only give him \eqn{70.31/0.782411 = 89.86} CAD back. He lost just
over 10% of his original 100 CAD!

But would Michael, had he declined the trade in 2020, only to accept in 2021,
feel even more foolish? He would have gotten 100 CAD for only 70.31 USD in 2020,
but now the trade cost him 78.24 USD. In this world, Michael paid over 11% more
in 2021 than he would have a year before... also a bad deal. But wait? Surely
Michael declining the trade must be *exactly as bad* for him as Jonathan's trade
was for Jonathan? *And it would have been!* Michael would have given Jonathan
7.93 USD more in 2021 than in 2020 for the same 100 CAD... since he would have
ultimately given Jonathan 78.24 USD, this is a loss (in 2021 dollars) of
\eqn{7.93/78.24 = 0.101}, i.e. the same 10.1% loss Jonathan would have
experienced.

Woof. This is confusing! We *can* make the math work out... but we could also
use better-behaved units!


<!-- Next introduce log scaling, nels, centinels, and then show the cool
compounding 1 % and linear change in centinels, ideally with the exhange rate
data. The below is from the blog post -->

# copied text, won't make sense from here
"Centinels" look like they could be nice units for staring at this, but what the
heck is 1 centinel? Amazingly, a centinel isn't that funky a unit. Plus one
centinel is pretty close to a 1% increase!

let's take a look:

```{r centinels_vs_percents}

x <- c(100, 5, pi)
# 1% increase
y <- x*1.01 
z <- x *0.99
w <- x / 1.01
centinel <- function(x, ref){
  100*log(x/ref)
}

percent <- function(x, ref){
  100*(x/ref -1)
}


percent(y, x)
# it's pretty close
centinel(y, x)


# compare to decreases
percent(z, x)
# these are all close too
percent(w, x)

centinel(z, x)
centinel(w, x)  

# percents go additively, though. So it's hard to think about compounding them
# which is what we usually want to do when things change

# if COVID went up by 1% a day for five consecutive days, it would go up by 
# about 5 %. 
1.01^5
# But if it went up by 1% a day for fifty consecutive days
1.01^50
# it would go up by more than 60%. This can really trip me up!

# centinels compound much more sensibly

compounded <- function(x, change, times){
  x*change^times
}

# compound 1% 50 times, you get a 64% increase
percent(compounded(x, 1.01, 50), x)

# but a 50 centinel increase :-)
centinel(compounded(x, 1.01, 50), x)

# try going backwards
percent(compounded(x, 0.99, 50), x)
# that's annoying. Only 39% now?!?!


centinel(compounded(x, 0.99, 50), x)
# still a change of 50 centinels

# and more precisely opposite 
centinel(compounded(x, 1/1.01, 50), x)

```
