---
title: "centiNels"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{centiNels}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
  , fig.width = 7
  , fig.height = 7
)
```

```{r setup, message=FALSE, include = F}
library(ratioScales)
library(tidyverse)
theme_set(theme_classic() +
          theme(axis.text.x = element_text(angle = 45,
                                           vjust = 1, hjust = 1)))
```


# Foolish percent differences

On 1 April, 2020, a Canadian dollar (CAD) cost 0.703 US dollar (USD). The
Canadian dollar gained strength pretty steadily for the next year, and by 1 July
it would cost 0.736 USD, by  1 October, 0.753 USD, by 4 January 2021, 0.782 USD,
and by April fools day 2021, 0.796 USD. How much is the exchange rate changing?

Often, such changes are described in terms of percent difference, even though
this is sort of insane. Below is a brief peek at the insanity of percentage
differences: any change in the exchange rate will be reflected as different
percentage changes, depending on which direction you're hoping to exchange
currency.

```{r percents are insane}
exch %>% 
  filter(date %in% as.Date(c("2020-04-01"
                     , "2020-07-01"
                     , "2020-10-01"
                     , "2021-01-04"
                     , "2021-04-01"))) %>% 
  mutate(pct_difference = round(100*(exRate_scale-1), 2)) %>% 
  select(date, direction, exRate, pct_difference) %>% 
  pivot_wider(id_cols = c("date")
              , names_from = "direction"
              , values_from = c("exRate", "pct_difference")
              , names_vary = "slowest"
              )
```

On 1 April 2021, US dollars had gotten much cheaper for Canadians than they had
been a year previously... one US dollar cost less than 90% as much as a it had a
year earlier (or a US dollar had lost just over 10% of its value in CAD). Amazingly,
over the same period, Canadian dollars got a bit over 11% more expensive for
Americans... the 10% gain corresponded to an 11% loss. 

To put this more concretely, if Jonathan traded 100 CAD with Michael for 70.31
USD on 1 April 2020, doubtless he'd feel the fool come 1 April, 2021, when
Michael would only give him $70.31/0.782411 = 89.86$ CAD back. He lost just
over 10% of his original 100 CAD!

But would Michael, had he declined the trade in 2020, only to accept in 2021,
feel even more foolish? He would have gotten 100 CAD for only 70.31 USD in 2020,
but now the trade cost him 78.24 USD. In this world, Michael paid over 11% more
in 2021 than he would have a year before... also a bad deal. But wait? Surely
Michael declining the trade must be *exactly as bad* for him as Jonathan's trade
was *good* for Jonathan? *And it would have been!* Michael would have given
Jonathan 7.93 USD more in 2021 than in 2020 for the same 100 CAD... since he
would have ultimately given Jonathan 78.24 USD, this is a loss (in 2021 dollars)
of $7.93/78.24 = 0.101$, i.e. the same 10.1% loss Jonathan would have
experienced.

Woof. This is confusing! We *can* make the math work out... but we could also
use a better-behaved scale... like the natural log scale.


# Nels instead 

Logarithmic transformation is an excellent solution\* to this problem. It makes
all ratio-based changes linear.  For now, 
let's go with it's "excellent." ,


*Familiarity and comfort with logarithms can be major drawbacks; we refer
users to the "divMult" vignette for further discussion and some good
workarounds*

The first step to making the math work here is rescaling the data by taking the
natural logarithm. This isn't revolutionary. One idea explored in
**ratioScales** is that we can simply use natural logarithm *units* as our
measure of change. Once log-transformed, changes in exchange rate in one
direction will be mirrored by changes in exchange rate in the other direction
*of the exact same magnitude*. **ratioScales** is all about plotting that
retains this, and in this vignette, we illustrate simply using a "base e" fold
change (one nel) as the plotting unit. This can be a sensible alternative to the
also sensible, common practice of, after a logarithmic transformation,
back-transforming data onto the original scale for plotting. **ggplot2** users
will likely be familiar with the functions `scale_*_log10()`, which do this.

Perhaps now, the idiomatic way to plot the currency data in **ggplot2** is to
use the raw exchange rates (we'll say, scaled to the starting value so we can
watch relative change), transform the y-axis by taking the base-10 logarithm,
but mark the axes by exponentiating back to the original exchange rate scale. 

Here is an example of the exchange rates on the original and base-10 logarithmic 
scales:

```{r idiomatic ggplot currency graphs,  out.width = '45%', out.height = '45%', fig.show = "hold"}

# first, without logarithmic rescaling
exch %>%
  ggplot(aes(date, exRate_scale, color = direction)) +
  geom_hline(yintercept = 1, width = 0.2 ) +
  geom_point() +
  geom_point(
    data = exch %>%
      filter(date %in% as.Date(c("2020-04-01"
                                 , "2020-07-01"
                                 , "2020-10-01"
                                 , "2021-01-04"
                                 , "2021-04-01")))
    , size = 5
    , shape = 23
    , color = "black"
    , fill = "red") +
  scale_color_manual(values = hcl.colors(4, "Plasma")[c(1,2)]) +
  labs(y = "exchange rate", main = "orignal scale")  +
  scale_y_continuous(n.breaks = 7)

# second, with log10 transformation of y-axis

exch %>% 
  ggplot(aes(date, exRate_scale, color = direction)) + 
  geom_hline(yintercept = 1, width = 0.2 ) +
  geom_point() +
  geom_point(
    data = exch %>% 
      filter(date %in% as.Date(c("2020-04-01"
                                 , "2020-07-01"
                                 , "2020-10-01"
                                 , "2021-01-04"
                                 , "2021-04-01")))
    , size = 5
    , shape = 23
    , fill = "red"
    , color = "black") +
  scale_color_manual(values = hcl.colors(4, "Plasma")[c(1,2)]) +
  labs(y = "exchange rate", main = "logarithmic scale") +
  scale_y_log10(n.breaks = 7)
```

The difference between these two graphs is pretty subtle. If you look closely,
you'll see the distance between 0.85 and 0.9 is substantially greater than the
distance between 1.15 and 1.2 on the logarithmic y-axis graph (second), whereas,
of course, they are the same distance apart on the arithmetic y-axis (first
graph). This gives us a feel for the fact that a decrease of 10% is a bigger
change than an increase of 10%, but it's still hard to see what is consistent
between the two directions of exchange (which above, we said was one reason to
prefer the log-transformations!).

An alternative is to become more comfortable with log-transformed data. Lots of
bases can make sense for logarithms (e.g., 2, 10, or e), here we
explore natural logarithms (base e = `exp(1)`). To do this, let's indulge some
nomenclature fun, then see what it can do for us.

First, let's get the most basic unit here... what do we call 1 natural log unit? One **N**atural **L**og unit is "one nel" (one NL; one "nel"). How big is one nel? Actually, kind of big: a fold change of over 2.5, so an increase of one nel is over 2.5x. 


<!-- Next introduce log scaling, nels, centinels, and then show the cool
compounding 1 % and linear change in centinels, ideally with the exhange rate
data. The below is from the blog post -->

# copied text, won't make sense from here
"Centinels" look like they could be nice units for staring at this, but what the
heck is 1 centinel? Amazingly, a centinel isn't that funky a unit. Plus one
centinel is pretty close to a 1% increase!

let's take a look:

```{r centinels_vs_percents}

x <- c(100, 5, pi)
# 1% increase
y <- x*1.01 
z <- x *0.99
w <- x / 1.01
centinel <- function(x, ref){
  100*log(x/ref)
}

percent <- function(x, ref){
  100*(x/ref -1)
}


percent(y, x)
# it's pretty close
centinel(y, x)


# compare to decreases
percent(z, x)
# these are all close too
percent(w, x)

centinel(z, x)
centinel(w, x)  

# percents go additively, though. So it's hard to think about compounding them
# which is what we usually want to do when things change

# if COVID went up by 1% a day for five consecutive days, it would go up by 
# about 5 %. 
1.01^5
# But if it went up by 1% a day for fifty consecutive days
1.01^50
# it would go up by more than 60%. This can really trip me up!

# centinels compound much more sensibly

compounded <- function(x, change, times){
  x*change^times
}

# compound 1% 50 times, you get a 64% increase
percent(compounded(x, 1.01, 50), x)

# but a 50 centinel increase :-)
centinel(compounded(x, 1.01, 50), x)

# try going backwards
percent(compounded(x, 0.99, 50), x)
# that's annoying. Only 39% now?!?!


centinel(compounded(x, 0.99, 50), x)
# still a change of 50 centinels

# and more precisely opposite 
centinel(compounded(x, 1/1.01, 50), x)

```
