---
title: "divMult and percDiff"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{divMult and percDiff}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
  , fig.width = 7
  , fig.height = 7
)
```

```{r setup, message=FALSE, include = F}
library(ratioScales)
library(dplyr)
library(tidyr)
theme_set(theme_classic() +
          theme(axis.text.x = element_text(angle = 45,
                                           vjust = 1, hjust = 1)))
```

<!-- this is a new vignette to illustrate use case for the divMult scale -->>
## Hook to illustrate a problem that needs to be solved (perhaps not exchange-rate focused)


## Introducing the divMult scale

<!-- generally want to show how existing scales challenge intuitions or engender misunderstanding-->

```{r idiomatic ggplot graphs,  out.width = '45%', out.height = '45%', fig.show = "hold"}

# first, without logarithmic rescaling

# second, with log10 transformation of y-axis


```
<!-- text explaining how divMult makes way more sense in many contexts -->
``` {r divMult plot}


```


<!-- Then, want to provide percDiff as a good compromise as well -->

# Percent differences can work too, if an appropriate scale is used

The problem with percent differences (as typically presented) is that they suggest
symmetry in confusing ways, and make the large relative changes near 0 and 100 
seem small. But, if an appropriate (i.e., logarithmic) scaling is used, familiar
percent differences make great axis ticks. 



```{r graphs showing percDiff scale vs. percent difference without rescaling}


```
