---
title: "gallery"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{gallery}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(ratioScales)
library(dplyr)
library(ggplot2)
```

# Gallery

This vignette illustrates the use of ratioScales in data visualizations on
topics ranging from exchange rates and college admissions to <!--everything else
-->

## Wealth and college admissions
In July 2023, Chetty et al. published [a
study](https://opportunityinsights.org/wp-content/uploads/2023/07/CollegeAdmissions_Paper.pdf)
that showed, among other things, that after controlling for SAT scores, wealthy
students had much higher attendance rates at the most elite US colleges. Their
data were visualized by Bhatia et al. in the [New York
Times](https://www.nytimes.com/interactive/2023/07/24/upshot/ivy-league-elite-college-admissions.html),
and later developed into an [interactive
toolkit](https://www.nytimes.com/interactive/2023/09/11/upshot/college-income-lookup.html)
that used the kinds of "divmult" labels we use in **ratioScales**, but they
plotted proportional differences linearly as in the original manuscript. Here,
we re-create some of their images, side by side, with and without ratio scales.

```{r admit}
admit %>% 
  filter(tier == "Ivy Plus", par_income_lab != "Top 1") %>% 
  ggplot(aes(par_income_lab
             , y =  rel_att_cond_app
             # , y = rel_attend
             # , y = rel_attend_sat
             , group = 1 )) +
  geom_hline(yintercept = 1) +
  stat_summary(fun = "mean"
               , geom = "line",  size = 1, color = "chocolate1") +
  stat_summary( fun.data = "mean_se"
                , geom = "pointrange", fatten  = 4, color = "chocolate") +
  labs(x = "parent income (percentile)"
       , y = "relative attendance rate, given application") + 
  # stat_summary(aes(y = rel_apply), geom = "line") +
  #  stat_summary(aes(y = rel_apply), geom = "pointrange") +
  # scale_y_ratio(tickVal = "divMult", slashStar = FALSE, n = 12) + 
  theme_classic()  

admit %>% 
  filter(tier == "Ivy Plus", par_income_lab != "Top 1") %>% 
  ggplot(aes(par_income_lab, rel_attend, group = 1 )) +
  geom_hline(yintercept = 1) +
  stat_summary(geom = "line", size = 1, color = "chocolate1") +
  stat_summary(geom = "pointrange",  fatten  = 4, color = "chocolate") + 
  # scale_x_sqrt() +
  # scale_x_log10() +
  theme_classic()  

admit %>% 
  filter(tier == "Ivy Plus", par_income_lab != "Top 1") %>% 
  ggplot(aes(par_income_lab, rel_attend, group = 1 )) +
  geom_hline(yintercept = 1) +
  stat_summary(geom = "line", size = 1, color = "chocolate1") +
  stat_summary(geom = "pointrange",  fatten  = 4, color = "chocolate") + 
  # scale_x_sqrt() +
  # scale_x_log10() +
  theme_classic()  
  

```

