---
title: "gallery"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{gallery}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(ratioScales)
library(dplyr)
library(ggplot2)
library(patchwork)
```
# Gallery
This vignette illustrates the use of ratioScales in data visualizations on
topics ranging from exchange rates and college admissions to <!--everything else
-->
## Wealth and college admissions
<!-- I think this is the right basic idea, but the data series to visualize
could be smarter. Overall, we want something that gives a nice clear picture
with the ratioScales. Also, simpler story (relative relative relative stuff is
infuriating) would be better. Keep exploring. --> 
In July 2023, Chetty et al. published [a
study](https://opportunityinsights.org/wp-content/uploads/2023/07/CollegeAdmissions_Paper.pdf)
that showed, among other things, that after controlling for SAT scores, wealthy
students had much higher attendance rates at the most elite US colleges. Their
data were visualized by Bhatia et al. in the [New York
Times](https://www.nytimes.com/interactive/2023/07/24/upshot/ivy-league-elite-college-admissions.html),
where the headline highlighted how at the most "Selective" colleges, parent
income boosted admissions rates after controlling for standardized test scores.
Their visualizations were later developed into an [interactive
toolkit](https://www.nytimes.com/interactive/2023/09/11/upshot/college-income-lookup.html)
Although the original Chetty et al. paper used traditional linear scales, the
NYT/upshot visualizations used the ratio scale we call "divMult"  in
**ratioScales**.

Here, we pulled out some additional relationships revealed by Chetty et al.'s
data, visualizing them, with and without ratio scales.

These data conflate admissions and "yield"... showing relative attendance rates
for students with parents in each income bracket after controlling (within
school tier/type) and relative application rates. A rate >1 may mean students
are admitted at higher rates, or that they choose to attend once offered
admission at higher rates, or a combination of both factors, compared to
expectations given by their test scores.
```{r admit, out.width = 100}
# create a base plot (linear scale)
bp <- admit %>%
  # filter(tier == "Ivy Plus", par_income_lab != "Top 1") %>%
  # not using: , "Highly selective private"
  filter(tier %in% c("Highly selective public", "Ivy Plus", "Other elite schools (public and private)")
         , par_income_lab != "Top 1") %>%
  group_by(tier, par_income_lab) %>% 
  summarize(mean_rel_att = mean(rel_att_cond_app)
            , sd_rel_att = sd(rel_att_cond_app)) %>% 
  ggplot(aes(par_income_lab
             , y =  mean_rel_att
             , color = tier
             , group = tier
             )) +
  geom_hline(yintercept = 1) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) + 
  labs(x = "parent income (percentile)"
       , y = "relative attendance rate \n(controlling for test score and application rate)") +
  theme_classic() +
  scale_y_continuous(breaks = c(0.5, 1/1.5, 1/1.25, 1, 1.25, 1.5, 2)
                     , limits = c(0.48, 2.5)
                     , labels = function(x){round(x,2)}
                     ) +
  scale_color_viridis_d()

# add ratio scale
ratioscaled <- bp +
  scale_y_ratio(tickVal = "divMult"
                , slashStar = FALSE
                , breaks = c(0.5, 1/1.5, 1/1.25, 1, 1.25, 1.5, 2)
                , limits = c(0.48, 2.5)
                ) +
   annotate(x = "80-90", y = 2, label = "divMult ratio scale", geom = "text") 

#print plots side by side
bp + 
  annotate(x = "80-90", y = 2, label = "traditional linear scale", geom = "text") +
  theme(legend.position = "none") + 
  ratioscaled +
  labs(y = NULL, color = NULL)

```

Of course, both plots show the same data, but they seem to tell somewhat
different stories. Both plots show that the wealthiest students attend the most
selective colleges... "Ivy Plus" and "Other elite," at much higher-than-expected
rates... over double for the top 0.1% at the Ivy Plus schools. A tier down, the
"Highly selective colleges" show an opposing pattern, where the wealthiest
students disproportionately do not attend. Presumably these students don't
receive an admissions disadvantage at these public schools, but instead are
choosing to attend those most selective colleges.

The ratio scale highlights, among the "Highly selective public" colleges, the
dramatically low attendance rate for the wealthiest students. This dramatic
difference is compressed on the linear scale.

## Home value change during great recession

Across the United States, home values plummeted during the great recession, but 
unevenly across regions. 

It is [common to describe
appreciation](https://www.investopedia.com/terms/a/appreciation.asp) as a
percentage difference relative to the starting value. Here, we visualize changes
in home values, where for each region the Zillow Home Value Index <!-- does this
need a trademark or something? --> is computed, the baseline value is the value
in the first reporting period.

```{r US home value, out.width = 100}

base_HV_plot <- ushs %>% 
  filter(RegionType == "msa", !is.na(TypicalHomeValue)) %>% 
  group_by(RegionID) %>% 
  mutate(NormTypicalHomeValue = TypicalHomeValue/TypicalHomeValue[1] ) %>% 
  ggplot(aes(RecordDate, NormTypicalHomeValue, color = RegionName)) +
  geom_line() + 
  facet_wrap(~StateName
             # , scales = "free_y"
             , ncol = 10
             ) +

  labs(y = "percent change in typical home value", x = "date") +
  theme_minimal() + 
  theme(legend.position = "none"
        , axis.text.x = element_text(angle = 90
                                     # , hjust =1
                                     , vjust = 0.5
                                     )) 

base_HV_plot + scale_y_ratio(tickVal = "percDiff"
                # , n = 6
                # , limits = c(0.5, 4.3)
                , breaks = c(0.5, 0.75, 1, 1.5, 2, 3.5)
                ) 

```

These pictures show that losing 50% of a value is as big a change as gaining
100% (and much bigger than gaining 50%). This also makes it easy to spot when
trends are synchronous and asynchronous in a short time period or across time
series with different starting points.


