---
title: "gallery"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{gallery}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(ratioScales)
```

# Gallery

This vignette illustrates the use of ratioScales in data visualizations on
topics ranging from exchange rates and college admissions to <!--everything else
-->

## Wealth and college admissions
In July 2023, Chetty et al. published [a
study](https://opportunityinsights.org/wp-content/uploads/2023/07/CollegeAdmissions_Paper.pdf)
that showed, among other things, that wealthy students had much higher
admissions rates among all colleges and universities, with many of the largest
gaps at the most selective schools. Their data were visualized by Bhatia et al.
in the [New York
Times](https://www.nytimes.com/interactive/2023/07/24/upshot/ivy-league-elite-college-admissions.html),
and later developed into an [interactive
toolkit](https://www.nytimes.com/interactive/2023/09/11/upshot/college-income-lookup.html)
that used ratio scales; the original Chetty et al. manuscript did not. Here, we
re-create some of their images, side by side, with and without ratio scales.

