---
title: "Ratio scales and centinels"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Ratio scales and centinels}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
  , fig.width = 7
  , fig.height = 7
)
```

```{r setup, message=FALSE}
library(ratioScales)
library(tidyverse)
theme_set(theme_classic() +
          theme(axis.text.x = element_text(angle = 45,
                                           vjust = 1, hjust = 1)))

```
## Hook

I've been staring at the NY Times COVID data obsessively, nearly daily, for most 
of the past two years. Not only am I on the look out (a *sentinel*), but I even
try to make some decisions on the basis of these data, trying to understand the 
disease prevalence in my region as I evaluate which activities seem like 
reasonable risks. 

I've noticed the NYT shows the case rates on an arithmetic scale, which I think
makes my work a bit harder. Here is an exploration of some alternative scales
and units (all the scale and unit ideas here are from [Jonathan
Dushoff](https://mac-theobio.github.io/dushoff.html), but I wrote this up in
March 2022 to understand them all), which might help me make more sense of the
data and think more clearly about the risks I take.

First, grab the data, which are freely available on github: 

## Data acquisition, crunching

```{r get_nyt_data}

# vid <- read.csv("https://github.com/nytimes/covid-19-data/raw/master/rolling-averages/us-states.csv")
data(vid, package = "ratioScales", verbose = TRUE)
head(vid)

```

The first thing I'm going to do is establish a reference point for myself. I 
gathered with >150 vaccinated people without masks in October of 2021. This was 
right at the edge of what seems like an ok risk to me, so I'll set it as a 
reference point. 

```{r reference data}

ref_date <- vid %>% filter(date =="2021-10-17") %>%  select(state, ref_case_rate = cases_avg_per_100k)

new_vid <- vid %>% left_join(ref_date, by = "state")
```

Just to make my life easier, I'm going to add a couple columns that rescale the
average daily case rate per 100,000 residents vis-a-vis this reference date, one
on the arithmetic scale and the other on the log scale. Also, to make life 
easier I'll drop all states except my imediate neighbors in Maryland.

```{r rescale data and filter states }
near_states <-c("Maryland", "Virginia", "Delaware", "District of Columbia", "West Virginia", "Pennsylvania")

nel_vid <- new_vid %>% mutate(nel_rate = log(cases_avg_per_100k/ref_case_rate)
                          , prop_rate = cases_avg_per_100k/ref_case_rate
                          , date = as.Date(date)) %>%
  filter(state %in% near_states) 
```

# Scales

I've grown accustomed to staring at the data on an arithmetic scale. Here, "1" 
represents a rate equivalent to the one on that date when I was somewhat 
comfortable with >150 other maskless, vaccinated people: 

```{r plot_arithmetic}
nel_vid %>%  
  ggplot(aes(date, prop_rate))+
  geom_point(size = 0.2)+
  facet_wrap(~state) +
  geom_hline(yintercept = 1, size = 0.1) +
  geom_vline(xintercept = as.Date("2021-10-17"), color = "red") +
  scale_x_date() +
  labs(y = "Cases relative to my reference", x = "date")

```

Now, "5" means the cases were 5x as high, and "0.2" (which is hard to pick out)
means the cases were 1/5 as high. The first thing that jumped out to me is that,
in all the neighboring states, cases are now lower than they were on my
reference date! But, on this scale, it's very hard to see by how much.

An alternative to using the arithmetic scale is to use a *natural*
logarithmic scale. Whereas in the graph above, 2 represented double my reference
case rate, on the natural log scale, 2 means... "*2 nels*". The really nice
thing is that "nels" are symmetrical in terms of changes in the case rate: if
the rate *doubles*, the graph goes up by about 0.7 nels, and if it *halves*, the
graphs go down by the same about (about 0.7 nels).

Let's look at the graph plotted in "nels": 
```{r first_nel_plot}
nel_vid %>%  
  ggplot(aes(date, nel_rate))+
  geom_point(size = 0.2)+
  facet_wrap(~state) +
  geom_vline(xintercept = as.Date("2021-10-17"), color = "red") +
  scale_x_date() +
  geom_hline(yintercept = 0, size = 0.1) +
  labs(x = "date", y = "Cases relative to my reference \n(nels)")
```

WOW! I see so many things in this graph! The first thing to jump out is how
incredibly good it got in summer 2021, which was very hard to discern in the
graph above. Looks like in early July last year, cases were *3 nels lower* than
even my reference, which I considered to be a relatively good time (though the
Delta variant was pretty horrid then, especially in some parts of the USA).

Another things that jumps out is that maybe I should drop the early data. First
off, there's lots of reasons to ignore them anyways, since testing
infrastructure was so different then, and second off, they make the range of the
graph go really low, below minus **nine** "nels."

So, what is a *nel*? You've probably guessed, it's called that because it is one
Natural Log unit, (one NL; one "nel"). A nel is a big number: 1 nel is over 2.5
times my reference rate, and 2 nels is over 7 times my reference rate. 

Once I drop the very early data, I might see that a "nel", like a
["bel"](https://en.wikipedia.org/wiki/Decibel), might be an awkwardly sized unit
for the kinds of changes I'm most interested in, so we'll look at "*centinels*"
instead. This is the same idea as changing from a proportion to a percent. 

# Centinels!!!

```{r nel_to_centinel}
# drop before april 2020
    
cut_vid <- nel_vid %>% filter(date > as.Date("2020-04-01"))

# we are used to using percents instead of proportions
cut_vid %>%  
  ggplot(aes(date, prop_rate))+
  geom_point(size = 0.2)+
  facet_wrap(~state) +
  geom_hline(yintercept = 1, size = 0.1) +
  geom_vline(xintercept = as.Date("2021-10-17"), color = "red") +
  scale_x_date() +
  scale_y_continuous(labels = scales::percent) +
  labs(y = "Cases relative to my reference", x = "date")

# now "centinels" instead of "nels"
cut_vid %>% 
  ggplot(aes(date, nel_rate*100))+
  geom_point(size = 0.2)+
  facet_wrap(~state) +
  geom_vline(xintercept = as.Date("2021-10-17"), color = "red") +
  scale_x_date() +
  geom_hline(yintercept = 0, size = 0.1) +
  labs(x = "date", y = "Cases relative to my reference \n(centinels)")
```

"Centinels" look like they could be nice units for staring at this, but what the
heck is 1 centinel? Amazingly, a centinel isn't that funky a unit. Plus one
centinel is pretty close to a 1% increase!

let's take a look:

```{r centinels_vs_percents}

x <- c(100, 5, pi)
# 1% increase
y <- x*1.01 
z <- x *0.99
w <- x / 1.01
centinel <- function(x, ref){
  100*log(x/ref)
}

percent <- function(x, ref){
  100*(x/ref -1)
}


percent(y, x)
# it's pretty close
centinel(y, x)


# compare to decreases
percent(z, x)
# these are all close too
percent(w, x)

centinel(z, x)
centinel(w, x)  

# percents go additively, though. So it's hard to think about compounding them
# which is what we usually want to do when things change

# if COVID went up by 1% a day for five consecutive days, it would go up by 
# about 5 %. 
1.01^5
# But if it went up by 1% a day for fifty consecutive days
1.01^50
# it would go up by more than 60%. This can really trip me up!

# centinels compound much more sensibly

compounded <- function(x, change, times){
  x*change^times
}

# compound 1% 50 times, you get a 64% increase
percent(compounded(x, 1.01, 50), x)

# but a 50 centinel increase :-)
centinel(compounded(x, 1.01, 50), x)

# try going backwards
percent(compounded(x, 0.99, 50), x)
# that's annoying. Only 39% now?!?!


centinel(compounded(x, 0.99, 50), x)
# still a change of 50 centinels

# and more precisely opposite 
centinel(compounded(x, 1/1.01, 50), x)

```

We like centinels. The name is fun, and they are helpful for thinking about
compounded changes. They balance around the reference point. Maybe we can use
them more. But it will take some practice to become fluent in interpreting them.

I rely on benchmarks: Beyond the 1% thing, I also know that 70 centinels is
roughly doubling, 140 is doubling ~2x, and 210 is doubling ~3x. To get
to know centinels better, though, I need to get more familiar with centinel
values on the "divMult scale:" instead of "69.3," "138.6," and "207.9," and
translating, I'll have an easier time when I see "2," "4," and "8."

## Ratio scales

```{r centinels_and_ratios}

cut_vid %>% 
  ggplot(aes(date, prop_rate))+
  geom_point(size = 0.2)+
  facet_wrap(~state) +
  geom_vline(xintercept = as.Date("2021-10-17"), color = "red") +
  scale_x_date() +
  scale_y_ratio(
   tickVal = "divMult"
    , slashStar = FALSE
    , sec.axis = sec_axis( trans = ~log(.)*100
                          , breaks = scales::breaks_extended(n = 6)
                          , name = "centinels")
 
    ) + 
  geom_hline(yintercept = 1, size = 0.1) +
  labs(x = "date", y = "Cases relative to my reference \n(divMult scale)" )

```

With the ratio scale as a guide, I feel I can navigate centinels more easily...
but maybe I just prefer the ratio scale? On the ratio scale, the reference is 1
(it could be $\times 1$  or $\div 1$, either is an identity). Then, labels
indicate symmetrical geometric changes (e.g., $\times 2, \div 2; \times 12, \div
12$). This scale preserves the symmetry of log scales, but the labels have
units, and operators, that make concrete, intuitive sense to me.

What have I learned about COVID-19 in my area? Things got really bad this
winter, which I knew... but I have a clearer quantitative sense of now. Also,
as I already knew, times have been getting better, but now I see how fast! If I
take the NYT daily case rate as a good prevalence estimate, the prevalence of
COVID-19 is now less than half what it was in mid-October 2021, when the Delta
surge was subsiding. Although mask mandates in the region have been lifted
recently, the prevalence continues to fall at a good clip, with only slight
indications that lifting mandates has slowed the subsidance. If my risk
tolerance is similar to what it was in the fall, I should feel pretty good now
about gathering with people who are vaccinated and important to me :-).

